Modelling the Lymphatic Metastatic Progression Pathways of OPSCC from Multi-Institutional Datasets

### 2022-12-09 src/tex/ms.tex
[showyourwork] overleaf sync

% Define document class
\documentclass[twocolumn]{aastex631}
\usepackage{showyourwork}
\usepackage{multirow}
\usepackage{hyperref}
\usepackage{cleveref}

@@

@@

When treating cancer, either with radiotherapy or via surgery, the aim is to irradiate or resect as much of the present malignancies as possible. This includes the primary tumor mass and metastases that are sufficiently large to be detected using in-vivo imaging modalities such as \gls{ct}, \gls{mri}, or \gls{pet}. But it also includes regions of possible microscopic spread, which these modalities cannot detect, to increase the patient's probability of cure \cite{poortmans_internal_2020,murthy_prostate-only_2021}. Microscopic disease can currently only be observed by a pathological examination of the tissue. Consequently, clinicians that are presented with cancer patients need to routinely assess the risk of microscopic involvement in parts of the body that appear unsuspicious.

Since in this work we consider \gls{hnscc}, which frequently spreads through the lymphatic system to form regional nodal metastases,

@@

\subsection*{Lymphatic Spread as Hidden Markov Model}
\label{subsec:previous_work:spread_as_hmm}

@@

The transition from one time-step to another is governed by the transition probability $P\left( \mathbf{X}[t+1]=\boldsymbol{\xi}_i \mid \mathbf{X}[t]=\boldsymbol{\xi}_j \right)$, which can conveniently be collected into a transition matrix when we enumerate all $2^V$ distinct possible states $\boldsymbol{\xi}_i$ with $i \in \left\{ 1,2, \ldots, 2^V \right\}$ of lymphatic involvement:
%
\begin{equation}
    \mathbf{A} = \left( A_{ij} \right) = \left( P\left( \mathbf{X}[t+1]=\boldsymbol{\xi}_i \mid \mathbf{X}[t]=\boldsymbol{\xi}_j \right) \right)
\end{equation}
%
The term $P\left( \boldsymbol{\xi}_i \mid \boldsymbol{\xi}_j \right)$ describes the probability to transition from the hidden state of lymphatic involvement $\boldsymbol{\xi}_j$ to the state $\boldsymbol{\xi}_i$ within between the time $t$ and $t+1$. Using a \gls{dag} as depicted in \cref{fig:graph_with_obs}, we can parameterize this transition probability in the following way:
%
\begin{equation}
    P\left( \boldsymbol{\xi}_i \mid \boldsymbol{\xi}_j \right) = \prod_{v \leq V} Q\left( \xi_{iv} ; \xi_{jv} \right) P \left( \xi_{iv} \mid \left\{ \xi_{jr} \right\}_{r \in \operatorname{pa}(v)} \right)^{1 - \xi_{jv}}
\end{equation}
%
In this equation, we have denoted \glspl{lnl} that are parents of \gls{lnl} $v$ with the symbol $r\in\operatorname{pa}(v)$. Also, $\xi_{iv}$ denotes the value that \gls{lnl} $v$ takes on when the patient is in state $\boldsymbol{\xi}_i$.
\begin{figure}
    \centering
    \includegraphics[width=0.6\linewidth]{figures/graph_with_obs.png}
    \caption{\Gls{dag} representing a possible abstraction of the lymphatic network comprising the tumor (red shaded circle) and \glspl{lnl} I through IV as hidden binary \glspl{rv} (blue outlined circles). Attached to each of these is the corresponding observed \gls{rv} (orange shaded squares). Lymphatic flow is depicted in the form of parameterized arrows (red and blue) that represent the probability of spread along the respective arc per time-step. Sensitivity and specificity (orange arrows) connect the hidden \glspl{rv} to the diagnosis.}
    \label{fig:graph_with_obs}
\end{figure}

@@

\subsection*{Parameterization of the Hidden Markov Model}
\label{subsec:previous_work:parametrization}

blabla

@@

\bibliography{bib}

>>>>

% Define document class
\documentclass[twocolumn]{aastex631}
\usepackage{showyourwork}
\usepackage{multirow}

%% References
\usepackage[%
    style=numeric-comp,%
    sorting=none,%
    maxnames=20,%
    url=false,%
]{biblatex}
\addbibresource{bib.bib}

@@

% Cross-referencing
\usepackage{hyperref}
\AtEndPreamble{\usepackage{cleveref}}

@@

When treating \gls{hnscc}, either with radiotherapy or via surgery, the aim is to irradiate or resect as much of the present malignancies as possible. This includes the primary tumor mass and metastases that are sufficiently large to be detected using in-vivo imaging modalities such as \gls{ct}, \gls{mri}, or \gls{pet}. But it also includes regions of possible microscopic spread, which these modalities cannot detect, to increase the patient's probability of cure \cite{poortmans_internal_2020,murthy_prostate-only_2021}. Microscopic disease can currently only be observed by a pathological examination of the tissue. Consequently, clinicians that are presented with cancer patients need to routinely assess the risk of microscopic involvement in parts of the body that appear clinically node negative.

This work concerns itself with \gls{opscc}, which commonly spreads lymphatically. The only two publicly available datasets on nodal involvement patterns in \gls{opscc}, at the time of writing, show that 80\% of patients where either cN+ or -- if available -- pN+ \cite{ludwig_dataset_2022}. Since the exact number and location of lymph nodes varies from patient to patient, the nodes are grouped into anatomically defined \glspl{lnl}. These levels are then often irradiated or resected prophylactically due to the risk of harboring occult metastases despite negative findings from imaging modalities. In current clinical practice, \gls{ctv-n} definition is instead mostly based on guidelines \cite{gregoire_ct-based_2003,gregoire_delineation_2014,gregoire_delineation_2018,eisbruch_intensity-modulated_2002,biau_selection_2019,chao_determination_2002,vorwerk_guidelines_2011,ferlito_elective_2009} that are derived from the observed prevalence of involvement in an \gls{lnl} for a given tumor location.

These guidelines, however, do not account for the personal risk of the patients that may depend greatly on their diagnosis. E.g., macroscopic metastases detected via \gls{pet} in the \glspl{lnl} II and III may increase the risk for occult disease in \gls{lnl} IV over the case of a patient who presents with a clinically N0 neck.

A model for estimating the risk of microscopic disease, given a personal diagnosis based on \glspl{bn} was previously developed \cite{pouymayou_bayesian_2019} and subsequently adapted to make use of \glspl{hmm} to include T-category in an intuitive manner \cite{ludwig_hidden_2021}. These models were trained on a dataset of early T-category (T1 and T2) patients that could be reconstructed from \cite{sanguineti_defining_2009}. Since this publication only includes details about the involvement of the ipsilateral \glspl{lnl} I, II, III, and IV, the models were limited to these nodal levels as well.

With the publication of larger and more detailed datasets \cite{ludwig_dataset_2022}, the risk estimation models may be extended as well to include the additionally reported \glspl{lnl} V and VII. This work aims to do that and at the same time demonstrate the capability 
of the previously introduced \gls{hmm} to adapt to real patient data from a multi-centric cohort of XXX \gls{opscc} patients.

@@

\subsection{Lymphatic Spread as Hidden Markov Model}
\label{subsec:previous_work:spread_as_hmm}

@@

The transition from one time-step to another is governed by the transition probability $P\left( \mathbf{X}[t+1]=\boldsymbol{\xi}_i \mid \mathbf{X}[t]=\boldsymbol{\xi}_j \right)$, which can conveniently be collected into a transition matrix when we enumerate all $2^V$ distinct possible states $\boldsymbol{\xi}_i$ with $i \in \left\{ 1,2, \ldots, 2^V \right\}$ of lymphatic involvement:
%
\begin{equation}
    \mathbf{A} = \left( A_{ij} \right) = \left( P\left( \mathbf{X}[t+1]=\boldsymbol{\xi}_i \mid \mathbf{X}[t]=\boldsymbol{\xi}_j \right) \right)
\end{equation}
%
The term $P\left( \boldsymbol{\xi}_i \mid \boldsymbol{\xi}_j \right)$ describes the probability to transition from the hidden state of lymphatic involvement $\boldsymbol{\xi}_j$ to the state $\boldsymbol{\xi}_i$ within between the time $t$ and $t+1$. Using a \gls{dag} as depicted in \cref{fig:graph_with_obs}, we can parameterize this transition probability in the following way:
%
\begin{equation}
    \label{eq:transition_prob}
    P\left( \boldsymbol{\xi}_i \mid \boldsymbol{\xi}_j \right) = \prod_{v \leq V} Q\left( \xi_{iv} ; \xi_{jv} \right) P \left( \xi_{iv} \mid \left\{ \xi_{jr} \right\}_{r \in \operatorname{pa}(v)} \right)^{1 - \xi_{jv}}
\end{equation}
%
In this equation, we have denoted \glspl{lnl} that are parents of \gls{lnl} $v$ with the symbol $r\in\operatorname{pa}(v)$. Also, $\xi_{iv}$ denotes the value that \gls{lnl} $v$ takes on when the patient is in state $\boldsymbol{\xi}_i$. Lastly, the term $Q(a;b) \in \{ 0,1 \}$ is there only to

\begin{figure}
    \centering
    \includegraphics[width=0.6\linewidth]{figures/graph_with_obs.png}
    \caption{\Gls{dag} representing a possible abstraction of the lymphatic network comprising the tumor (red shaded circle) and \glspl{lnl} I through IV as hidden binary \glspl{rv} (blue outlined circles). Attached to each of these is the corresponding observed \gls{rv} (orange shaded squares). Lymphatic flow is depicted in the form of parameterized arrows (red and blue) that represent the probability of spread along the respective arc per time-step. Sensitivity and specificity (orange arrows) connect the hidden \glspl{rv} to the diagnosis.}
    \label{fig:graph_with_obs}
\end{figure}

@@

\subsection{Parameterization of the Hidden Markov Model}
\label{subsec:previous_work:parametrization}

The term $P \big( X_v = \xi_{iv} \mid \left\{ X_r = \xi_{jr} \right\}_{r \in \operatorname{pa}(v)} \big)$ in \cref{eq:transition_prob} implicitly depends on how we parameterize the arcs of the underlying \gls{dag}. If we use the graph displayed in \cref{fig:graph_with_obs}), we can write down a conditional probability table for the probability of $X_3$ being healthy or metastatic, based on the state of the parent \gls{lnl} $\operatorname{pa}(3) = 2$:

\begin{table}[h!]
    \centering
    \begin{tabular}{|cc|cc|}
        \hline
        & & \multicolumn{2}{c|}{$X_2$} \\
        & & 0 & 1 \\
        \hline
        \multirow{2}{*}{$X_3$} & 0 & $1 - b_3$ & $(1 - b_3)(1 - t_{23})$ \\
        & 1 & $b_3$ & $1 - b_3 - t_{23} + b_3 t_{23}$ \\
        \hline
    \end{tabular}
\end{table}

@@

\printbibliography[%
    title={References},%
]
